---
title: "Analysis"
author: "pg611@scarletmail.rutgers.edu"
date: "2024-03-28"
output: html_document
---

```{r}
library(psych)
library(readr)
## Warning: package 'readr' was built under R version 4.3.2
library(factoextra)

library(FactoMineR)
## Warning: package 'FactoMineR' was built under R version 4.3.3
library(magrittr)
library(NbClust)
library(readr)
library(lattice)
library(ggplot2)
library(ggridges)
library(ggvis)
library(ggthemes)
library(cowplot)
library(gapminder)
library(gganimate)
library(dplyr)
library(tidyverse)
library(grid)
library(gridExtra)
library(RColorBrewer)
library(cluster)
library(readr)
library(factoextra)
library(magrittr)
library(NbClust)
library(ggplot2) # for visualization
library(reshape2) # for data manipulation
library(factoextra)
library(FactoMineR)
library(psych)
library(corrplot)
library(devtools)

```


# DATASET DESCRIPTION :

The dataset comprises 22 rows and 10 columns, representing the social media usage of students in a class. Each row corresponds to a specific student, while each column provides information about the usage of different social media platforms and a variable indicating whether the student experiences trouble falling asleep.


# DATA DICTIONARY

Here's the data dictionary for the dataset:

ID: Unique identifier for each student.

Instagram Usage: The amount of time spent by each student on Instagram.

LinkedIn Usage: The amount of time spent by each student on LinkedIn.

Snapchat Usage: The amount of time spent by each student on Snapchat.

Twitter Usage: The amount of time spent by each student on Twitter.

Whatsapp Usage: The amount of time spent by each student on WhatsApp.

Youtube Usage: The amount of time spent by each student on YouTube.

OTT (Over-the-top media service) Usage: The amount of time spent by each
student on OTT platforms.

Reddit Usage: The amount of time spent by each student on Reddit.

Trouble_falling_asleep: A variable indicating whether the student experiences trouble falling asleep. It may have values like "Yes" or "No".










```{r}

sc_dataset <- read.csv("/Users/parul/OneDrive/Desktop/MVA/Midterm_Final.csv", row.names=1)


```



The correlation matrix provides insight into the relationships between different variables. In this case, the correlation matrix displays the correlations between different social media usage measurements (LinkedIn, Snapchat, Twitter, WhatsApp, Youtube, OTT, Reddit).

```{r}

#Get the Correlations between the measurements

correlation_matrix <- cor(sc_dataset[,1:8][-1])
correlation_matrix

```

•	A correlation coefficient close to 1 indicates a strong positive relationship, while a coefficient close to -1 indicates a strong negative relationship.

•	A correlation coefficient close to 0 indicates no linear relationship between the variables.

```{r}
# Visualize correlation matrix
ggplot(melt(correlation_matrix), aes(Var1, Var2, fill = value)) +
  geom_tile(color = "white") +
  scale_fill_gradient2(low = "blue", high = "red", mid = "white", 
                       midpoint = 0, limit = c(-1,1), space = "Lab", 
                       name="Correlation") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, size = 10, hjust = 1)) +
  coord_fixed()

```


For Example :

LinkedIn.Usage vs. Snapchat.Usage: A moderate positive correlation (0.3254), suggesting that as LinkedIn usage increases, Snapchat usage tends to increase as well, but not strongly.

LinkedIn.Usage vs. Twitter.Usage: A weak negative correlation (-0.0934), indicating a slight negative relationship between LinkedIn and Twitter usage.

LinkedIn.Usage vs. Whatsapp.Usage: A moderate positive correlation (0.2258), indicating a moderate tendency for LinkedIn and WhatsApp usage to increase together.
 and so on..
 

These correlations provide insights into the relationships between different social media usage variables, which can be further explored through principal component analysis (PCA) to identify underlying patterns and reduce the dimensionality of the data.


# PCA


•	PCA helps in reducing the dimensionality of the dataset while retaining most of the variability present in the original data.

•	It identifies patterns and relationships between the different social media usage metrics and potentially identifies underlying factors influencing these patterns.

•	The results suggest that the first few principal components explain a significant portion of the variability in social media usage behaviors, indicating potential underlying factors driving these behaviors.

```{r}
sc_dataset_pca <- prcomp(sc_dataset[,1:8][,-1],scale=TRUE) 
sc_dataset_pca
summary(sc_dataset_pca)
```

•	The standard deviations represent the variability captured by each principal component (PC).

•	PC1 has the highest standard deviation (1.4211), followed by PC2 (1.3815), indicating that PC1 explains the most variability in the data, followed closely by PC2.







The column "Trouble_falling_asleep" likely represents a categorical variable indicating whether individuals in the dataset experience trouble falling asleep.


```{r}
library(factoextra)



# Eigenvalues are sdev^2
(eigen_sc_dataset <- sc_dataset_pca$sdev^2)
names(eigen_sc_dataset) <- paste("PC",1:3,sep="")
eigen_sc_dataset
sumlambdas <- sum(eigen_sc_dataset)
sumlambdas
propvar <- eigen_sc_dataset/sumlambdas
propvar
cumvar_sc_dataset <- cumsum(propvar)
cumvar_sc_dataset
matlambdas <- rbind(eigen_sc_dataset,propvar,cumvar_sc_dataset)
rownames(matlambdas) <- c("Eigenvalues","Prop. variance","Cum. prop. variance")
round(matlambdas,4)
summary(sc_dataset_pca)
sc_dataset_pca$rotation
print(sc_dataset_pca)
sc_dataset_pca$x


sc_datasettyp_pca <- cbind(data.frame(sc_dataset$Trouble_falling_asleep),sc_dataset_pca$x)
sc_datasettyp_pca



# Means of scores 
tabmeansPC <- aggregate(sc_datasettyp_pca[,2:4],by=list(Trouble_falling_asleep=sc_dataset$Trouble_falling_asleep),mean)
tabmeansPC
tabmeansPC <- tabmeansPC[rev(order(tabmeansPC$Trouble_falling_asleep)),]
tabmeansPC
tabfmeans <- t(tabmeansPC[,-1])
tabfmeans
colnames(tabfmeans) <- t(as.vector(tabmeansPC[1]$Trouble_falling_asleep))
tabfmeans





# Standard deviations of scores 
tabsdsPC <- aggregate(sc_datasettyp_pca[,2:4],by=list(Trouble_falling_asleep=sc_dataset$Trouble_falling_asleep),sd)
tabfsds <- t(tabsdsPC[,-1])
colnames(tabfsds) <- t(as.vector(tabsdsPC[1]$Trouble_falling_asleep))
tabfsds
t.test(PC1~sc_dataset$Trouble_falling_asleep,data=sc_datasettyp_pca)
t.test(PC2~sc_dataset$Trouble_falling_asleep,data=sc_datasettyp_pca)
t.test(PC3~sc_dataset$Trouble_falling_asleep,data=sc_datasettyp_pca)




## F ratio tests
var.test(PC1~sc_dataset$Trouble_falling_asleep,data=sc_datasettyp_pca)
var.test(PC2~sc_dataset$Trouble_falling_asleep,data=sc_datasettyp_pca)
var.test(PC3~sc_dataset$Trouble_falling_asleep,data=sc_datasettyp_pca)




# Levene's tests (one-sided)
library(car)
(LTPC1 <- leveneTest(PC1~sc_dataset$Trouble_falling_asleep,data=sc_datasettyp_pca))
(p_PC1_1sided <- LTPC1[[3]][1]/2)
(LTPC2 <- leveneTest(PC2~sc_dataset$Trouble_falling_asleep,data=sc_datasettyp_pca))
(p_PC2_1sided=LTPC2[[3]][1]/2)
(LTPC3 <- leveneTest(PC3~sc_dataset$Trouble_falling_asleep,data=sc_datasettyp_pca))
(p_PC3_1sided <- LTPC3[[3]][1]/2)
```

T-tests:

T-tests have been performed to compare the means of the principal component scores between groups based on the variable "Trouble_falling_asleep" (Yes/No).

For PC1 and PC2, the p-values are greater than 0.05, suggesting that there is no significant difference in the means between the groups.

However, for PC3, the p-value is less than 0.05, indicating a significant difference in the means between the groups.
Variance Homogeneity Tests:

Levene's tests have been conducted to assess the homogeneity of variance for each principal component between the two groups.

For PC1 and PC2, the p-values are greater than 0.05, indicating that there is no significant difference in variance between the groups.

Similarly, for PC3, the p-value is also greater than 0.05, suggesting no significant difference in variance between the groups.






```{r}
# Different PCA Method. 
res.pca <- PCA(sc_dataset[,1:8][,-1], graph = FALSE)
print(res.pca)

fviz_eig(sc_dataset_pca, addlabels = TRUE)

plot(eigen_sc_dataset, xlab = "Component number", ylab = "Component variance", type = "l", main = "Scree diagram")
plot(log(eigen_sc_dataset), xlab = "Component number",ylab = "log(Component variance)", type="l",main = "Log(eigenvalue) diagram")

```


•	PC1 accounts for 28.85% of the total variance, PC2 for 27.26%, and subsequent components account for decreasing proportions of variance.

•	The cumulative proportion of variance explained by the principal components indicates how much of the total variance in the dataset is captured by each successive component. 

•	In this case, the first two components capture over 56% of the total variance, while the first three components capture over 70.5%.


```{r}
fviz_pca_var(sc_dataset_pca,col.var = "cos2",
             gradient.cols = c("#FFCC00", "#CC9933", "#660033", "#330033"),
             repel = TRUE)
```



This visualization helps interpret how different social media variables contribute to the principal components identified through PCA, with the color indicating the strength of their contribution and the position in the scatterplot revealing their relationships with each other and with the principal components.

For example in this case, Twitter, LinkedIn, and WhatsApp appear darker in the visualization compared to the other variables, it implies that these three variables have higher squared cosines, indicating that they contribute more significantly to the principal components identified through PCA.


```{r}
fviz_pca_ind(sc_dataset_pca, col.ind = "cos2", 
             gradient.cols = c("#FFCC00", "#CC9933", "#660033", "#330033"), 
             repel = TRUE)
```


Each dot in the scatter plot represents a single individual.
The position of a dot on the horizontal axis (Dim1) and the vertical axis (Dim2) reflects their scores on the two main categories, or principal components, identified by PCA.


Dim1 : A higher score on this axis (upper right) indicate users who engage more with professional networking platforms (LinkedIn), news or informational content across platforms, or educational resources.

Dim2 : A higher score on this axis (upper left or lower right) might suggest users who lean towards entertainment-focused platforms like OTT services, Whatsapp, social media for casual interaction (Snapchat), or platforms known for humor or memes (Reddit).

```{r}

biplot(sc_dataset_pca)

```

This combines above two plots into one.


```{r}
fviz_pca_biplot(res.pca, repel = TRUE,
                col.var = "#FC4E07", 
                )

```

```{r}
fviz_pca_ind(res.pca,
             geom.ind = "point", # show points only (nbut not "text")
             col.ind = sc_dataset$Trouble_falling_asleep, # color by groups
             palette = c("#00AFBB", "#E7B800", "#FC4E07"),
             addEllipses = TRUE, # Concentration ellipses
             legend.title = "Groups"
)
```


This visualization helps to understand how individuals with different levels of "Trouble falling asleep" are distributed in the principal component space, potentially revealing any patterns or associations between social media usage and sleep-related issues.



```{r}
fviz_pca_biplot(res.pca, 
                # Individuals
                geom.ind = "point",
                fill.ind = sc_dataset$Trouble_falling_asleep, col.ind = "black",
                pointshape = 21, pointsize = 2,
                palette = "jco",
                addEllipses = TRUE,
                # Variables
                alpha.var ="contrib", col.var = "contrib",
                gradient.cols = "RdYlBu",
                
                legend.title = list(fill = "Trouble Sleeping", color = "Contrib",
                                    alpha = "Contrib")
)




```


This biplot visualizes a PCA analysis where the data points represent individuals and the variables might be related to sleep or sleep troubles. The coloring and transparency of the points and variable shapes will likely help identify which variables are most highly correlated with the principal components (Dim1 and Dim2) and how sleep troubles are positioned relative to those components.







While PCA identifies linear combinations of variables that explain the maximum amount of variance, FA goes a step further by attempting to identify the underlying latent factors that explain the observed correlations between variables. This can provide more interpretable results by uncovering the theoretical constructs that drive the observed patterns in the data.



# FACTOR ANALYSIS 


It reveals the underlying structure of users' behavior across various social media platforms such as Instagram, LinkedIn, Snapchat, Twitter, WhatsApp, YouTube, OTT, and Reddit.

```{r}
#FACTOR ANALYSIS

fit.pc <- principal(sc_dataset[,1:8][-1], nfactors=4, rotate="varimax") 
fit.pc
round(fit.pc$values, 3)
fit.pc$loadings

# Communalities
fit.pc$communality
fit.pc$scores
# Play with FA utilities

```


Variables with higher absolute loadings on a particular factor are more strongly associated with that factor.
For instance, Twitter.Usage has a high loading on the first factor (RC1), suggesting that this factor represents usage patterns related to Twitter.
Similarly, OTT has high loadings on multiple factors, indicating its association with different underlying constructs.



```{r}

fa.parallel(sc_dataset[,1:8][-1]) # See factor recommendation
fa.plot(fit.pc) # See Correlations within Factors
fa.diagram(fit.pc) # Visualize the relationship
vss(sc_dataset[,1:8][-1]) # See Factor recommendations for a simple structure


```


Factor analysis revealed four latent factors (RC1, RC2, RC3, and RC4) derived from the social media dataset.

•	RC1 represents high-usage social media platforms conducive to extensive content consumption and social interaction, encompassing Twitter, WhatsApp, and OTT services. 

•	RC2 reflects platforms oriented towards learning and professional networking, exemplified by YouTube and LinkedIn.

•	RC3 captures Snapchat's distinctive role, primarily focused on instant image sharing and communication. 

•	RC4 highlights Reddit, known for its community-based discussions and content sharing. 



# CLUSTER ANALYSIS

Cluster analysis of the provided social media dataset enables the identification of distinct user segments based on their usage patterns across various platforms, facilitating a deeper understanding of user behavior and preferences.
```{r}
#CLUSTER ANALYSIS

matstd.sc_dataset <- scale(sc_dataset[,1:8])
sc_dataset_scaled <- scale(sc_dataset[,1:8])

# Creating a (Euclidean) distance matrix of the standardized data 
dist.sc_dataset <- dist(matstd.sc_dataset, method="euclidean")

fviz_nbclust(matstd.sc_dataset, kmeans, method ="wss")+geom_vline(xintercept = 4, linetype = 2)

```

•	By selecting four clusters, we aim to strike a balance between maximizing the homogeneity within clusters while minimizing the complexity of the segmentation. 

•	With four clusters, we can effectively group users with similar usage patterns and behaviors.


```{r}

# Hierarchial Clusiering
res.hc <- matstd.sc_dataset %>% scale() %>% dist(method = "euclidean") %>%
  hclust(method = "ward.D2")

# Lets see what the optimal numbers of clusers are
# Compute

res.nbclust <- sc_dataset[,1:8] %>% scale() %>% NbClust(distance = "euclidean", min.nc = 3, max.nc = 13, method = "complete", index ="all") 


fviz_dend(res.hc, k = 4, # Cut in four groups
          cex = 0.5, # label size
          k_colors = c("#2E9FDF", "#00AFBB", "#E7B800", "#FC4E07"),
          color_labels_by_k = TRUE, # color labels by groups
          rect = TRUE # Add rectangle around groups
)

```

•	There are around 20 items or observations in the dataset.

•	These items have been grouped into approximately 4 clusters.

•	The two items labeled “ki567” and “ak2001” are most similar to each other in the dataset

•	The items labeled “MVA37@S” and “Harvey” are least similar to each other in the dataset according to the criteria used to create the dendrogram.


```{r}

set.seed(123)
km.res <- kmeans(matstd.sc_dataset, 3, nstart = 25)
# Visualize
fviz_cluster(km.res, data = matstd.sc_dataset, ellipse.type = "convex",palette = "jco",ggtheme = theme_minimal())

```

•	The visualization of these clusters using fviz_cluster reveals the grouping of users into different segments characterized by their respective social media usage behaviors. 

•	Each cluster, represented by ellipses, showcases the concentration of users with similar usage tendencies within the feature space defined by the standardized social media usage variables.




```{r}

# If your data has outliears , use PAM method
pam.res <- pam(matstd.sc_dataset, 3)
# Visualize
fviz_cluster(pam.res)


```


•	The Partitioning Around Medoids (PAM) clustering method is particularly suitable for datasets with outliers as it employs representative points (medoids) instead of means for cluster assignment.

•	In this context, the PAM algorithm was applied to the standardized social media dataset to create three clusters based on the medoids. 

•	As shown, 15801 and masinl are the outliers in our dataset.





# MY DATA COMPARED TO OTHERS:


Z-scores allow us to understand how each student's social media usage compares to the overall distribution of usage for each platform. Positive z-scores indicate values above the mean, while negative z-scores indicate values below the mean.

```{r}
social_media_data1 <- read.csv("/Users/parul/OneDrive/Desktop/MVA/Midterm_Final.csv")

# Select the columns containing social media usage data
social_media_data1 <- sc_dataset[,1:8][, -1]  # Exclude the first column 

# Calculate mean and standard deviation for each social media variable
means <- colMeans(social_media_data1)
std_devs <- apply(social_media_data1, 2, sd)

# Calculate z-scores for each student
z_scores <- scale(social_media_data1, center = means, scale=std_devs)

print(z_scores)
```

For me - "drphy", the z-scores for various social media platforms are as follows:

LinkedIn Usage: 1.38331213

Snapchat Usage: -0.52468968

Twitter Usage: -0.4297799

Whatsapp Usage: 1.79843283

Youtube Usage: 2.33688179

OTT: 0.194410495

Reddit: -0.3230682


Interpretation:

This indicates that I have above-average usage for LinkedIn, Whatsapp, Youtube, and OTT (Over-the-Top) platforms, as their z-scores are positive, indicating values above the mean.
On the other hand, have below-average usage for Snapchat, Twitter, and Reddit, as their z-scores are negative, indicating values below the mean.

These z-scores provide a standardized measure for my social media usage relative to the mean usage of each platform among all students in the dataset.


# TAKE AWAY 

In the class, students can be categorized into two distinct groups based on their social media usage patterns. One group comprises students who spend significantly more time on various social media apps, while the other group consists of students who spend comparatively less time on these apps.

Additionally, there are a few outliers within these groups, indicating students who exhibit extreme behavior by spending disproportionately more time on specific types of apps compared to their peers.
The main lesson learned from this analysis is the need to decrease social media usage due to its potential impact on sleep and other aspects of our well-being. Below, I've outlined insights into how each platform can affect our sleep.

```{r}
library(dplyr)  # For data manipulation

# Load the dataset
social_media_data <- read.csv("/Users/parul/OneDrive/Desktop/MVA/Midterm_Final.csv")
# Remove irrelevant columns if any
social_media_data <- social_media_data[, -1]  # Assuming the first column is ID and not needed

# Check for missing values and handle them if necessary
# social_media_data <- na.omit(social_media_data)

# Separate data for individuals having trouble sleeping and those not having trouble sleeping
trouble_sleeping <- social_media_data %>% filter(Trouble_falling_asleep == "Yes")
no_trouble_sleeping <- social_media_data %>% filter(Trouble_falling_asleep == "No")

# Calculate the variance for each group
variance_trouble_sleeping <- apply(trouble_sleeping[, -9], 2, var)
variance_no_trouble_sleeping <- apply(no_trouble_sleeping[, -9], 2, var)

# Compare variances
comparison <- data.frame(
  Variable = colnames(social_media_data)[1:8],  # Assuming first 8 columns are variables
  Variance_Trouble_Sleeping = variance_trouble_sleeping,
  Variance_No_Trouble_Sleeping = variance_no_trouble_sleeping
)

# Print the comparison
print(comparison)
```

From the comparison results:

Instagram Usage: The variance of Instagram usage is higher among individuals having trouble sleeping compared to those not having trouble sleeping.

LinkedIn Usage: Similarly, the variance of LinkedIn usage is higher among individuals having trouble sleeping compared to those not having trouble sleeping.

Snapchat Usage: The variance of Snapchat usage is relatively similar between individuals having trouble sleeping and those not having trouble sleeping.

Twitter Usage: Interestingly, the variance of Twitter usage is much higher among individuals not having trouble sleeping compared to those having trouble sleeping.

WhatsApp Usage: The variance of WhatsApp usage is higher among individuals having trouble sleeping compared to those not having trouble sleeping.

Youtube Usage: The variance of Youtube usage is higher among individuals not having trouble sleeping compared to those having trouble sleeping.

OTT (Over-the-top media services): The variance of OTT usage is relatively similar between individuals having trouble sleeping and those not having trouble sleeping.

Reddit: The variance of Reddit usage is higher among individuals having trouble sleeping compared to those not having trouble sleeping.
